Translation of B Implementations to the LLVM: Specification

### 2014-07-07 DOCUMENTATION/DerivedTypes.tex
Small updates about array

Two approaches are possible to support arrays and structures. An approach is 
importing reusable libraries of AtelierB that are reported in the official 
document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}. 
This approach is supported, but the generated code is more complex and it needs 
more intermediate steps to manipulate the data of imported components. 
Moreover, this approach allows dynamic allocation for arrays, but it is 
strongly not recommended for safety-critical systems. A second approach enjoys 
the direct correspondence between B0 native types and LLVM and this approach 
is presented here. This section describes the translation of derived data 
types showed in section~\ref{sec:data}.

@@

Initializing a record with a sequence of expressions:
\[
  \textcolor{gray}{accountRecord :=}  \underbrace{rec(\underbrace{name:xx’name, balance:(xx’balance - amt)}_{seq\ Exp})}_{Record}
\]

Setting a record field to a variable:
\[
  \textcolor{gray}{xx :=}  \underbrace{accountRecord'balance}_{RecField} 
\]

@@

\begin{pascalcode}
xx := rec(id:1, balance:10);
xx'balance := xx'balance-1;
\end{pascalcode}

The lines 1 and 2 are represented in translation in LLVM code:

@@

The $RecField$ rule gets an element value from the record. The same rule is 
used on the left side of attribution as a $LValue$ and right side of 
attribution as an $Expression$. The functions \SBPos(n) and \SBType(n)  get 
respectively the position and type of $n$ element in the structure.

\begin{align*}
\begin{split}
  \trad{n}{\B{RecField}} \isdef \\
   &\LET \ell = \lbl  \AND  t = n.type \AND v = \B{head}(n.element) \AND\\
   &code \BE \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \IN \\
   &\IF \B{n.base.type} \mbox{ is \B{BOOL or NAT or INT} \AND n.element.length=1}\\
   &\quad \THEN  code \sep  \ell  \sep \SBType(\B{n})\\
   &\IF \B{n.base.type} \mbox{ is \B{RecField} }  \THEN \\
   &\LET \ell2 = \lbl  \AND  t = n.type \AND v = \B{tail}(n.element) \AND \\
   &\quad n2.base  \BE ?? \AND n2.element = \B{head}(n.element) \AND a, b, c = \trad{n2}{RecField}
\IN\\
   &\quad  code \conc  \llvm{"\PH{\ell2} = getelementptr \PH{t}* \PH{\ell}, i32 0, i32  \SBPos(\B{n})\nl"} \\
   &\quad \sep  \ell2  \sep \SBType(\B{n})\\
\end{split}
\end{align*}

The attribution from record structure to a record structure variable, line 1 
from last B code example, is translated by a sequence of application of rule 
$RecField$. The nested structures will be designed in the next project phase.
%TODO after test the first examples.

@@

%[[ Write a text here to define the rule for it]]

There are three attributions cases related to array.
\begin{description}

\item[Case 1.1] - Initializing an unique value to an array:
\[
\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..2 \times 0..2 \times 0..3  }_{seq\ Exp} \times \underbrace{\{5\}}_{Exp} }_{Product}\}}_{Array}
\]
\item[Case 1.2] - Initializing different values to an array\footnote{B0 definition requires the declaration of all values.}:
\[
\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(\underbrace{(0 \mapsto 0) \mapsto  0)}_{seq Exp} \mapsto \underbrace{5}_{Exp} , ... ,
(\underbrace{(2 \mapsto 2) \mapsto  3)}_{seq\ Exp} \mapsto \underbrace{5}_{Exp} }_{seq\ Map}\}}_{Array}
\]

Theses cases use a declared constant that is attributed to the variable:
\begin{llvmcode}
@array = global [3 x [3 x [4 x i32]]] 
[[3 x [4 x i32]] 
	[[4 x i32] [i32 5, i32 5, i32 5, i32 5], 
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]], 
 [3 x [4 x i32]] 
	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5],
 	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]],
 [3 x [4 x i32]] 
	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]]], align 16
\end{llvmcode}

\begin{align*}
\begin{split}
  \trad{array}{\B{Type}} \isdef \\
   & \LET \quad \IF \B{array} \mbox{ is \B{Product} } \THEN \\
   & \quad \quad  dec = \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Product}}\\
   & \quad \ELSIF array \mbox{ is \B{Map} } \THEN\\
   & \quad \quad  dec = \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Map}} \\
   & \quad \IN \llvm{" global \PH{dec}"}\\
\end{split}
\end{align*}
\VGM{Alguma sugestão com relação as funções:head, tail.}

\begin{align*}
\begin{split}
  \trad{type, seqIntl, value}{\B{Product}} \isdef \\
   &\LET \B{h = head(seqIntl)} \AND\\
   & \B{times = h.end - h.start + 1} \AND\\
   & values = \trad{times,type,value}{\B{TV}} \IN\\
   & \IF \B{seqIntl.length} == 1 \THEN  \\
   & \llvm{"[\PH{times} x \PH{type}] [\PH{values}]"}  \\
   & \ELSE \LET \\
   &  tvalues = \trad{type,tail(seqIntl),values}{\B{Product}} \\
   & \IN \\
   & \llvm{"[\PH{times} x \PH{type}]  \PH{tvalues}"}  \\
\end{split}
\end{align*}

\begin{align*}
\begin{split}
  \trad{times, type, value}{\B{TV}} \isdef \\
   &\LET res1 = \llvm{" \PH{type} \PH{value} "} \IN\\
   & \IF times == 1 \THEN \llvm{"\PH{res1}"} \\
   & \ELSE \LET   res2 =   \trad{times-1,type,value}{\B{TV}}  \IN  \\
   & \llvm{"\PH{res1} , \PH{res2}"} \\
\end{split}
\end{align*}

The ``head'' function returns the first element from sequence.
The ``tail'' function returns the given sequence without its first element.

@@

\item[Case 2.1] - Updating a selected element from array by a value of variable:
\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
\]
\item[Case 2.2] - Setting a selected element from array to a variable:
\[ 
\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem}
\]
\item[Case 3] - Attribution between arrays\footnote{B0 definition requires the same type.}:
\[
\underbrace{arr1}_{Array} := \underbrace{arr2}_{Array} 
\]
\end{description}

@@

The arrays in B0 are represented by functions where an index points to an 
element. By default, the element of index not initialized has a zero value.
The different ways to evaluate an array are described as follow:

@@

An alternative way to represent these attributions cannot be using sequential attributions as follow:

@@

It is not possible, because B array attribution are atomic. The B array 
attribution are made in two steps. First, the new value is calculated and
copied to a temporal place. Second, the sequence of new values are setted 
in theirs address. Finally, considering these steps and the execution of
following code then the value of ``array2(0,0,1)'' is ``2''.

@@

\begin{llvmcode}
%7 = getelementptr [2 x [2 x [4 x i32]]]* @Array2, i32 0, i64 1, i64 1, i64 3
\end{llvmcode}

@@

This example is translated in four steps. First, the variables \%x, \%y, \%z 
are allocated. Second, the respective values 0, 1, 3 are stored in variables. 
Third, the variables are loaded to local variables and converted to 64 bits. 
Finally, the address of array is calculated in local variable \%7 and the value 
7 is stored in it.  
\begin{figure}[h]
\begin{llvmcode}
%xx = alloca i32
%yy = alloca i32
%zz = alloca i32
store i32 1, i32* %xx
store i32 1, i32* %yy
store i32 3, i32* %zz
%1 = load i32* %zz
%2 = sext i32 %1 to i64
%3 = load i32* %yy
%4 = sext i32 %3 to i64
%5 = load i32* %xx
%6 = sext i32 %5 to i64
%7 = getelementptr[2x[2x[4xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7         ;setting an element
\end{llvmcode}
\end{figure}

@@

\paragraph{Product} declares all address and a unique value from array, it is represented by \textbf{case 1.1}.

@@

@@

The element access of array can need a shift in indexes. When the domain of 
array is an interval and it is not started from zero, it must adjust the 
indexes shifting the access, because the addressing  of array in intermediate 
representation LLVM starts at 0. To access an element from array the formula is 
$address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$. The following rule 
\footnote{The function $TSymbol$ returns the ``start'', ``end'' and type of the one 
index.} makes the shift for each index of array. \VGM{TODO: The function TSymbol
returns the ``start'', ``end'' and type of the one index.}

@@

\paragraph{Relevant translation restrictions:}
\begin{itemize}
\item  B0 definition - An array must be initialized declaring all values 
to avoid no determinism.
\item B0 definition from CommenC - The operator overwrite from arrays is not supported, for example:

>>>>

Two approaches are possible to support arrays and structures. One approach is 
importing reusable libraries of AtelierB that are reported in the official 
document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}. 
This approach is supported, but the generated code is more complex and it needs 
more intermediate steps to manipulate the data of imported components. 
Moreover, this approach allows dynamic allocation for arrays, but it is 
strongly not recommended for safety-critical systems. A second approach enjoys 
the direct correspondence between B0 native types and LLVM and this approach 
is presented here. This section describes the translation of derived data 
types showed in Section~\ref{sec:data}.

@@

\begin{description}
\item[Case 1.1] Initializing a record with a sequence of expressions:
\[
  \textcolor{gray}{accountRecord :=}  \underbrace{rec(\underbrace{name:xx’name, balance:(xx’balance - amt)}_{seq\ Exp})}_{Record}
\]

\item[Case 1.2] Setting a record field to a variable:
\[
  \textcolor{gray}{xx :=}  \underbrace{accountRecord'balance}_{RecField} 
\]

\item[Case 2]  Attribution between structures:
\[
  \underbrace{newRecord}_{Record} :=  \underbrace{accountRecord}_{Record} 
\]
\end{description}

@@

\begin{pascalcode}
xx := rec(id:1, balance:10);
xx`balance := xx`balance-1;
\end{pascalcode}

Lines 1 and 2 are represented in translation in LLVM code:

@@

The $RecField$ rule gets an element value from the record. The same rule is 
used on the left side of attribution as an $LValue$ and right side of 
attribution as an $Expression$. The functions \SBPos(n) and \SBType(n)  get 
respectively the position and type of $n$ element in the structure.

\begin{align*}
\begin{split}
  \trad{n}{\B{RecField}} \isdef \\
   &\LET \ell = \lbl  \AND  t = n.type \AND v = \B{head}(n.element) \AND\\
   &code1 \BE \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \IN \\
   &\IF \B{n.element} \mbox{ is \B{BOOL or NAT or INT} \AND n.element.length=1}\\
   &\quad \THEN  code1 \sep  \ell  \sep \SBType(\B{n})\\
   &\IF \B{head(n.element)} \mbox{ is \B{RecField} }  \THEN \\
   &\LET n2 \BE new \B{RecField}() \AND  \B{n2.base}  \BE \B{head}(n.element)\\
   &\AND \B{n2.element} \BE \B{tail}(n.element) \AND \\
   &\quad code2, label, pos = \trad{n2}{RecField} \IN\\
   &\quad \llvm{"\PH{code1} \PH{code2}"} \sep  label  \sep pos \\
\end{split}
\end{align*}

The attribution from record structure to a record structure variable, Line 1 
from the last B code example, is translated by a sequence of applications of
the  rule $RecField$. This rule was also designed to support nested structures.

The structure attribution are made in atomic way and it is simple example:

\begin{pascalcode}
simpleRec := rec(elem1 : 2, elem2:3);
simpleRec := rec(elem1 : simpleRec`elem1+1, elem2: simpleRec`elem1+1);
\end{pascalcode}

This code is translated to the following code and solved in three steps: solve
the expression, store it to a temporal variable and  store the result in the
final place.

\begin{pascalcode}
VAR tmp1, tmp2 IN
tmp1 := 1;
tmp2 := 2;
simpleRec`elem1 := tmp1;
simpleRec`elem2 := tmp2;
END;
VAR tmp3, tmp4 IN
tmp3 := simpleRec`elem1 + 1;
tmp4 := simpleRec`elem1 + 1;
simpleRec`elem1 := tmp3;
simpleRec`elem2 := tmp4;
END
\end{pascalcode}

In this example, note that the final value of $simpleRec'elem2$ is 2. This 
happens due to atomic assignment. This translation is represented by the 
$Record$ rule and it is omitted, because it is meta rule very simple.

The \textbf{case 2} is the atrribution between records:

\begin{pascalcode}
  newSimpleRec = simpleRec;
\end{pascalcode}

The following code is its translation to LLVM:

\begin{llvmcode}
  %4 = bitcast %struct.database* %newSimpleRec to i8*
  %5 = bitcast %struct.database* %simpleRec to i8*
  call void @llvm.memcpy.p0i8.p0i8.i64(i8* %4, i8* %5, i64 12, i32 0, i1 false)
....
; Function Attrs: nounwind
declare void @llvm.memcpy.p0i8.p0i8.i64(i8* nocapture, i8* nocapture readonly, i64, i32, i1) #1
\end{llvmcode}

The instruction bitcast is used to get the reference pointer from source address 
and the destination address. The function \textit{memcpy} get a sequence of argument.
The first argument is a pointer to the destination, the second is a pointer to the source.
The third argument is an integer argument specifying the number of bytes to copy, the
fourth argument is the alignment of the source and destination locations, and the fifth
is a boolean indicating a volatile access. By default, the last two arguments are 0 and false.

@@

There are three attribution cases related to array.
\begin{description}

\item[Case 1.1] - Initializing a unique value to an array using Cartesian product:
\[
\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..2 \times 0..2 \times 0..3  }_{seq\ Exp} \times \underbrace{\{5\}}_{Exp} }_{Product}\}}_{Array}
\]

\item[Case 1.2] - Initializing different values to an 
array\footnote{B0 definition requires the declaration of all values.} using a sequence of map:
\[
\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(\underbrace{(0 \mapsto 0) \mapsto  0)}_{seq Exp} \mapsto \underbrace{1}_{Exp} , ... ,
(\underbrace{(2 \mapsto 2) \mapsto  3)}_{seq\ Exp} \mapsto \underbrace{27}_{Exp} }_{seq\ Map}\}}_{Array}
\]

@@

\item[Case 2.1] - Updating a selected element from array by a variable value :
\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
\]
\item[Case 2.2] - Setting a selected element from array to a variable:
\[ 
\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem}
\]
\item[Case 3] - Attribution between arrays\footnote{B0 definition requires the same type.}:
\[
\underbrace{arr1}_{Array} := \underbrace{arr2}_{Array} 
\]
\end{description}

@@

The arrays in B0 are represented by functions where an index points to an 
element. The different ways to evaluate an array are described as follows:

@@

An alternative way to represent these attributions cannot be by
using sequential attributions as follow:

@@

It is not possible because B array attributions are atomic. The B array 
attribution is made in two steps. First, the sequence of new value is calculated and
copied to its temporal place. Second, the sequence of new values is setted 
in its address. Finally, considering these steps and the execution of
following code then the value of ``array2(0,0,1)'' is ``2''.

@@

\begin{llvmcode}
%7 = getelementptr [2 x [2 x [4 x i32]]]* @array2, i32 0, i64 1, i64 1, i64 3
\end{llvmcode}

@@

This example is translated in four steps. First, the variables \%x, \%y, \%z 
are allocated. Second, the respective values 0, 1, 3 are stored in variables. 
Third, the variables are loaded to local variables and converted to 64 bits. 
Finally, the address of array is calculated in local variable \%7 and the value 
7 is stored in it.

\begin{llvmcode}
%xx = alloca i32
%yy = alloca i32
%zz = alloca i32
store i32 1, i32* %xx
store i32 1, i32* %yy
store i32 3, i32* %zz
%1 = load i32* %zz
%2 = sext i32 %1 to i64
%3 = load i32* %yy
%4 = sext i32 %3 to i64
%5 = load i32* %xx
%6 = sext i32 %5 to i64
%7 = getelementptr[2x[2x[4xi32]]]*@array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7         ;setting an element
\end{llvmcode}

@@

\paragraph{Product} declares all address and a unique value from array, it is represented by \textbf{case 1.1}.
The first case uses a declared constant that is attributed to the variable:
\begin{llvmcode}
@array = global [3 x [3 x [4 x i32]]] 
[[3 x [4 x i32]] 
	[[4 x i32] [i32 5, i32 5, i32 5, i32 5], 
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]], 
 [3 x [4 x i32]] 
	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5],
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]],
 [3 x [4 x i32]] 
	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]]], align 16
\end{llvmcode}

\begin{align*}
\begin{split}
  \trad{array}{\B{Type}} \isdef \\
   & \quad \IF \B{array} \mbox{ is \B{Product} } \THEN \\
   & \quad \quad \llvm{" global \PH{\trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Product}}}"}\\
%   & \quad \ELSIF array \mbox{ is \B{seq Map} } \THEN\\
%   & \quad \quad  \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Map}} \\
\end{split}
\end{align*}

\begin{align*}
\begin{split}
  \trad{type, seqSet, value}{\B{Product}} \isdef \\
   & \LET\\
   & \quad \B{times} \BE \B{head}(seqSet).end - \B{head}(seqSet).start + 1\\
   & \quad left \BE \trad{seqSet,type}{ArrayType} \IN \\
   & \IF \B{seqSet.length} == 1 \THEN  \\
   & \quad \LET right = \trad{times,type,value}{\B{TV}} \IN\\
   & \quad \llvm{"\PH{left} [\PH{right}]"}  \\
   & \ELSE \\
   & \quad \llvm{"\PH{left} \PH{\trad{type,tail(seqSet),value}{\B{Product}}}"}\\
\end{split}
\end{align*}

\begin{align*}
\begin{split}
  \trad{times, type, value}{\B{TV}} \isdef \\
   &\LET res1 \BE \llvm{" \PH{type} \PH{value} "} \IN\\
   & \IF times == 1 \THEN \llvm{"\PH{res1}"} \\
   & \ELSE\\
   & \quad \llvm{"\PH{res1} , \PH{\trad{times-1,type,value}{\B{TV}}}"} \\
\end{split}
\end{align*}

@@

The attribution using an sequence of Map are made in atomic way and it is simple example.

\begin{pascalcode}
 array2 := ( (0..1) * (0..1) * (0..3) * {1}); // Cartesian product
 array2 := { 0|->0|->0|->2, 0|->0|->1|->(array2(0,0,0)+1), ...}; // sequence of map
\end{pascalcode}

This code is translated in the following code and solved in three steps: solve 
the expression, store it to a temporal variable and  store the result in the 
final place.

\begin{pascalcode}
VAR tmp1, tmp2 IN
tmp1 := 2;
tmp2 := array(0,0,0)+1;
array(0,0,0) := tmp1;
array(0,0,1) := tmp2;
END;
\end{pascalcode}

In this example, note that the final value of $array(0,0,1)$ is 2. This 
happens due to atomic assignment. This translation is represented by the 
$Map$ rule and it is omitted, because it is meta rule very simple.

@@

The element access of array can need a shift in indexes. When the domain of 
array is an interval and it is not started from zero, it must adjust the 
indexes shifting the access, because the addressing  of array in intermediate 
representation LLVM starts at 0. To access an element from array the formula is 
$address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$. The function $TSymbol$
returns the ``start'', ``end'' and type of the one index. The following rule 
makes the shift for each index of array.

@@

The \textbf{case 3} is the atrribution between arrays. This translation is 
very similar to atribution between records.

\begin{pascalcode}
	newArray = simpleArray;
\end{pascalcode}

The following code is its translation to LLVM:

\begin{llvmcode}
%x4  = bitcast [10 x i32]* @newArray to i8*
%x5  = bitcast [10 x i32]* @simpleArray to i8*
call void @llvm.memcpy.p0i8.p0i8.i64(i8* %x4, i8* %x5, i64 40, i32 0, i1 false)
\end{llvmcode}

\paragraph{Relevant translation restrictions in B2LLVM:}
\begin{itemize}
\item  B0 definition - An array must be initialized declaring all values 
to avoid no determinism.
\item B0 definition from CommenC - The operator overwrite from arrays is not supported, for example:

### 2014-07-07 DOCUMENTATION/specification.tex
Small updates about array

@@

The data types \textbf{structures} and \textbf{arrays} require detailed 
representation in the abstract syntax structure from the table \ref{tab:node-attr}.

The section \ref{sec:StructArray} also presents other examples and rules for these derived data types.

@@

The rules to declare the structure type are:
\begin{align*}
\trad{Structure}{\B{Type}} \isdef \LET tps \BE \trad{Structure.elements}{\ListOf{\B{TField}}} \AND\\
 l \BE \local(\B{Structure.elements})  \IN  \llvm{"\PH{l}= type \{\PH{tps}\}\nl"}
\end{align*}
$$\trad{t}{\B{TField}} \isdef \LET res \BE  \trad{t.type}{\B{Type}} \IN \llvm{"\PH{res},"}$$

@@

[[ TODO: Write about the initialization (Structure and Array) unsing the concept of module/instance]]

The current specification of the code generator does not handle deferred sets.
These data types and logic/math operators compose expressions that is in 
the next section.

@@

\begin{align*}
\begin{split}
  \trad{a}{\B{Beq}} \isdef
  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Exp}} \IN \\
  & \llvm{"\PH{l}} \\
  & \llvm{\PH{r}} \\
  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
\end{split}
\end{align*}

@@

\textcolor{blue}{ The expressions have elements represented by pointers that need a special 
care. The attributions between these elements are made  directly by the memory 
address.  However, the expression resolution are made by parts, where the 
partial results are stored to temporary variables. Therefore, the expression 
resolution with pointers needs an additional instruction $LOAD$ for store the 
partial results. This translation is represented by $ValueItem$.}

\begin{align*}
\begin{split}
  \trad{n}{\B{ValueItem}} \isdef \\
   & \LET \ell = \lbl  \IN\\
   & \IF \B{n} \mbox{ is \B{RecField}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{RecField}} \\
   & \ELSIF \B{n} \mbox{ is \B{ArrayItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{ArrayItem}} \\
   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \\
   & \sep  \ell  \sep t'\\
\end{split}
\end{align*}

>>>>

There are also functions to manipulate the sequences: the $\B{``head''}$ function 
returns the first element from sequence; the $\B{``tail''}$ function returns 
the given sequence without its first element.

@@

The data types \textbf{structures} and \textbf{arrays} require detailed 
representation in the abstract syntax structure from the Table \ref{tab:node-attr}.

Section \ref{sec:StructArray} also presents other examples and rules for these derived data types.

@@

The rules to declare the structure type are:
\begin{align*}
\trad{Structure}{\B{Type}} \isdef \LET tps \BE \trad{Structure.elements}{\ListOf{\B{Field}}}\\
%\AND l \BE \local(\B{Structure.elements})  \IN  \llvm{"\PH{l}= type \{\PH{tps}\}\nl"}
  \\IN  \llvm{" type \{\PH{tps}\}\nl"}
\end{align*}
$$\trad{t}{\B{Field}} \isdef \LET res \BE  \trad{t.type}{\B{Type}} \IN \llvm{"\PH{res},"}$$

@@

%TODO
\VGM{[[ TODO: Write about the initialization (Structure and Array) unsing the concept of module/instance]]}

The current specification of the code generator does not handle deferred sets.
These data types and logic/math operators compose expressions that are explained
in the next section.

@@

\begin{align*}
\begin{split}
  \trad{a}{\B{Beq}} \isdef
  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{ValueItem}} \IN \\
  & \llvm{"\PH{l}} \\
  & \llvm{\PH{r}} \\
  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
\end{split}
\end{align*}

@@

\textcolor{blue}{The expressions have elements represented by pointers that need a special 
care. The attributions between these elements are made  directly by the memory 
address.  However, the expression resolution are made by parts, where the 
partial results are stored to temporary variables. Therefore, the expression 
resolution with pointers needs an additional instruction $LOAD$ for store the 
partial results. This translation is represented by $ValueItem$.}

\begin{align*}
\begin{split}
  \trad{n}{\B{ValueItem}} \isdef \\
   & \LET \ell \BE \lbl  \IN\\
   & \IF \B{n} \mbox{ is \B{RecField}} \THEN \LET p' \sep v' \sep t' = \trad{\B{n}}{\B{RecField}} \IN \\
   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \sep  \ell  \sep t'\\
   & \ELSIF \B{n} \mbox{ is \B{Record}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{Record}} \\
   & \ELSIF \B{n} \mbox{ is \B{ArrayItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{ArrayItem}} \\
   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \sep  \ell  \sep t'\\
   & \ELSIF \B{n} \mbox{ is \B{Array}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{Array}} \\
   & code \sep  \ell  \sep t'\\
   & \ELSE  p' \sep v' \sep t' = \trad{\B{n}}{\B{Exp}}
\end{split}
\end{align*}
